---
title: "EDA_Project"
author: "Rahman Rahimi"
date: "04 April 2018"
output: 
  html_document:
    toc: TRUE

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', include = TRUE, tidy = TRUE)
```

---
Lending Club Loan Data by Rahman Rahimi
========================================================
------
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.
# install.packages("rmarkdown")
# install.packages("ggplot2", dependencies = T) 
# install.packages("dplyr", dependencies = T)
# install.packages("GGally", dependencies = T)
# install.packages('memisc', dependencies = T)
# install.packages('corrplot', dependencies = T)

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

```

```{r echo=FALSE, Univariate_Plots}



suppressMessages(library(reshape2))
suppressMessages(library(reshape2))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(maps))
suppressMessages(library(RColorBrewer))
suppressMessages(library(GGally))
suppressMessages(library(scales))
suppressMessages(library(memisc))
suppressMessages(library(grid))
suppressMessages(library(gridExtra))
suppressMessages(library(corrplot))
suppressMessages(library(lubridate))
```


This dataset contains complete loan data for all loans issued through 2007-2015,
including the current loan status (Current, Late, Fully Paid, etc.)
and latest payment information. The file is a matrix of about 890 thousand 
observations and 75 variables. A data dictionary is provided in a separate file.


```{r results = 'hide', Load_the_Data}
# Load the Data
loan <- read.csv('loan.csv')
head(loan)
```

```{r}
tble <- tbl_df(loan)
```


# Univariate Plots Section

In this section, we perform some preliminary exploration of our dataset. we run 
some summaries of the data and create univariate plots to understand the
structure of the individual variables in our dataset.


```{r}
glimpse(tble)
```


```{r results='markup'}
df<- tble[ , colSums(is.na(tble)) < length(tble)/2]
head(df)
```

We first remove the columns that more than 50% of values are NaNs. As you can 
see from the table above, we have 50 columns remaining. Let's focus on these 
columns over the course of our EDA.


```{r results='hide'}
factor_vars<- df[,sapply(df, is.factor)]
head(factor_vars)
```


```{r results='hide'}
numerical_vars<- df[,sapply(df, is.numeric)]
head(numerical_vars)
```

Now, I have devide the entire dataset into 2 seperate ones, one for numeric
columns and one for factor columns. The tables above show a head of each of these
2 newly created dataframes.

```{r  results='markup'}
p1<- ggplot(data = df, aes(loan_amnt))+
  geom_histogram(binwidth = 1000)
p2<- ggplot(data = df, aes(loan_amnt))+
  geom_density(fill= 'gray')
grid.arrange(p1, p2, ncol = 2)
```

> Above, we can see the distribution of Loan Amount using 2 different plots.

```{r results='markup'}
summary(df$loan_amnt)
```

As we can see from the table above, the loan amount varies between as little as
$500 up to maximum of $35000. The mean of the loan amount is $14755 and as the 
boxplot suggests, the majority of the loans are somewhere between $8000 - $20000.
There are a large number of loans of $35000 as well.


```{r results='markup'}
loan_stat_df<- subset(df, !is.na(df$loan_status)) %>%
  group_by(loan_status) %>%
  summarise(Number=n())
loan_stat_df
```

Let us have a look at the Loans' Status. About 600 thousand loans are still
current. About 200 thousand are fully paid and about 45 thousand loans are
charged off. 


```{r results='markup'}
dataset.new <- filter(df, !loan_status %in% 
                        c("Does not meet the credit policy. Status:Charged Off",
                          "Does not meet the credit policy. Status:Fully Paid",
                      "Default"))
ggplot(data = dataset.new, aes(loan_status))+
  geom_bar(position = 'dodge')+ 
  labs(x = "Loan Status", title = "Distribution of Loan Status")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


# Univariate Analysis

* **What is/are the main feature(s) of interest in your dataset?**

A number of columns with Factor type and a the rest are continous or discrete
form of numerical values.
Among the numeric fields, the Loan Amount, Annual Income, Interest Rate are of
particular interest.Of all the categorical fields (Factors), Home ownership,
Loan Status, Loan Grade, Term, Verification Status are interesting.


* **Did you create any new variables from existing variables in the dataset?**

I have also added a ratio of Loan Amount / Annual Income at the end of the 
DataFrame, This might give us some insight on what ratio is the most likely or
most common range.

* **Of the features you investigated, were there any unusual distributions? Did
you perform any operations on the data to tidy, adjust, or change the form of
the data? If so, why did you do this?**

As we mentioned before, there are a number of columns with so many NaNs that 
could not be helpfu, so we decided to remove them.


# Bivariate and Multivariate Plots Section

* **Based on what you saw in the univariate plots, what relationships between
variables might be interesting to look at in this section?**

* I was particularly interested in the relationship between `loan amount` and the 
following fields:
+ Home Ownership
+ Employment Length
+ Loan Grade
+ Loan Status
+ Purpose
+ Interest Rate
+ Annula Income
+ Income-to-Loan-ratio
+ Verification Status
Below, let us take a deeper look at these relationships:

```{r results='markup', Bivariate_Plots}
p1 <- ggplot(data = df, aes(loan_amnt, color = grade))+
  geom_histogram(binwidth = 1000)+
  facet_grid(grade ~ .)
p2 <- ggplot(data = df, aes(loan_amnt, color = grade, fill = grade))+
  geom_density(binwidth = 1000)+
  facet_grid(grade ~ .)
grid.arrange(p1, p2, ncol = 2)
```

First, we look at the distribution of `Loan Amount` by Loan `Grade`. As you can see
majority of the loans are in `Grades A, B, and C`. The interesting observation
here was that as we move from grade A to Grade G, the loan amount distribution
becomes more homogeneous.

```{r results='markup'}
l1 <- ggplot(data = subset(df, !is.na(int_rate)), aes(int_rate, color = grade))+
  geom_histogram(bins = 50)+
  facet_grid(grade ~ .)+
  labs(title = "Interest Rate by Loan Grade")
l2 <- ggplot(data = subset(df, !is.na(int_rate)), aes(int_rate, color = grade))+
  geom_density(fill = 'SlateGray')+
  facet_grid(grade ~ .)+
  labs(title = "Interest Rate by Loan Grade")
grid.arrange(l1, l2, ncol = 2) 
```

Now, the distribution of `interest rates` by `Grade` is also presented above, as 
you can see, the lowest interest rates apply to Grade A. So the more costly loans
are in Grade G, F, E. 

I am so curious to know to whome these good loans (Grade A) are granted and 
what kind of obvious difference could be detected between good grades and bad
grades loan holders.


```{r results='markup'}
ggplot(data = subset(df, !is.na(annual_inc)), aes(x= annual_inc, y= id))+
  geom_point(alpha = 1/10, position = position_jitter(h=0))+
  labs(title = "Scatter Plot of Annual income by id")

```
The Scatter plot below shows the `self reported annual income` of all applicants. 
Obviously there are so many outliers that could be due to inputting error or some
other reasons. 

## finding outliers

```{r results='markup'}
annual_inc.na <- subset(df, !is.na(annual_inc))

q3<- quantile(annual_inc.na$annual_inc, 0.75)
q1<- quantile(annual_inc.na$annual_inc, 0.25)
iqr.annual_inc <- (q3 - q1)
range <- (q3 + 1.5 * iqr.annual_inc)


annual_inc_new <- annual_inc.na %>%
  filter(annual_inc < range)
```

We calculated the `IQR (Interquartile Range)` and decided to remove the outliers 
from the analysis.we use the common measure of 1.5 X IQR beyond 3rd quartile as
the outliers.


```{r results='markup'}
summary(annual_inc_new$annual_inc)
```
now we have removed the outliers and show the summary of the new data set above:


```{r results='markup'}
l1<- ggplot(data = annual_inc_new, aes(annual_inc))+
  geom_histogram(binwidth = 5000)
l2<- ggplot(data = annual_inc_new, aes(annual_inc))+
  geom_density(fill= 'gray')
l3<- ggplot(data = annual_inc_new, aes(y = annual_inc, x = 'id' ))+
  geom_violin(scale= 'area')
l4<- ggplot(data = annual_inc_new, aes(y = annual_inc, x = 'id' ))+
  geom_boxplot(scale= 'area')

grid.arrange(l1, l2, l3, l4, nrow = 2, ncol = 2)
```

Having removed the outliers, now we drew the same plots for Annual Income to
get a better picture of the income of the customers.

## Creating New Variables
**we are going to create a new variable called `Income_to_loan_ratio` below:**

```{r}

annual_inc_new$Income_to_loan_ratio <- annual_inc_new$annual_inc/annual_inc_new$loan_amnt
```


```{r results='markup'}
ggplot(data = annual_inc_new, aes(Income_to_loan_ratio))+
  geom_density(fill= 'gray')+
  coord_cartesian(xlim = c(0, 20))+
  labs(title = "Distribution (Density Plot) of the ratio of Annual Income to Loan Amount")

```

so the majority of people have a self_reported annula income of approximately
5 times the amount of loans they have been granted. 


```{r results='markup'}
ggplot(data = subset(df, ! home_ownership %in% c("ANY", "NONE", "OTHER")),
       aes(y = home_ownership , purpose))+
  geom_count(color = "Navy")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

**Let us have a look at the Home Ownership and Purpose of loans.**

The plot above, shows the number of loans grouped by the Home Ownership and
the purpose of the loan. Debt Consolidation is by far the most common purpose
of loans. But the porportion is different among different types of home ownership.
For example, the vast majority of Mortgage holders need loans for Debt 
Consolidation and credit cards. People who own their homes either do not need 
or do not apply for a loan through this website. As you can see, it is very
uncommon for the home owners to get loans to pay their credit cards debt compared
to the ones who either rent or paying mortgage.


```{r results='markup'}
ggplot(data = annual_inc_new, aes(y = annual_inc, x = id, fill = grade, 
                                  color = grade))+
  geom_boxplot(alpha = 1/3)+
  facet_grid(. ~ grade)+ 
  labs(title = "Distribution (Box Plot) of Annual income by Loan Grade")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
Box plots of Annual Income by Loan Grade are shown above.Looking at the median
line of the box plots, we see a downtrend from Grade A to D and then a slight
uptrend towards G.The Grade D has the narrowest range among all other grades.

## Home Ownership

```{r results='markup'}
df %>%
  filter(!is.na(home_ownership))%>%
  filter(! home_ownership %in% c("OTHER", "NONE", "ANY"))%>%
  group_by(home_ownership)%>%
  summarise(mean_amnt_loan = mean(loan_amnt),
            mean_annual_inc = mean(annual_inc),
            n=n())%>%
  ungroup()%>%
  arrange(desc(n))
```

Table above shows the average amount of loans and average annual income by 
different Home Ownership status. More than 443 thousand loans were granted to
those with a mortgaged home, but only 87 thosand to those who own their home.
A clear difference between the annual inmcome of those with different home
ownership is also shown.


```{r results='markup'}
ggplot(data = subset(df, (!is.na(home_ownership) & (! home_ownership %in% c("ANY", "NONE", "OTHER")))),
       aes(home_ownership, fill = grade,color = grade))+
  geom_bar()+ labs(title = "Distribution of Home Ownership by Loan Grade")
```

Here I brought the plot of Home Ownership by Loan Grade as a stacked bar chart.
The majority of loans are granted to those with Mortgage and grade B and C are
the most common ones.

## purpose of loan

```{r results='markup'}
df %>%
  filter(!is.na(purpose))%>%
  group_by(purpose)%>%
  summarise(mean_annual_inc = mean(annual_inc),
            mean_amnt_loan = mean(loan_amnt),
            n=n())%>%
  ungroup()%>%
  arrange(desc(n))
```

The table above, depicted the Average amount of loan and average annual income
based on the purpose of the loan. As we can see more than 520 thousand loans were
granted for debt consolidation and more than 200 thousand for credit card. For 
those with highest average annual income ($90271/year), home improvment is the
most common purpose, while those who have got loans for vacation have the lowest
average annual income!


```{r results='markup'}
ggplot(data = subset(df, !is.na(purpose)),
       aes(purpose, fill = loan_status,color = loan_status))+
  geom_bar()+ labs(title = "Distribution of Loan purpose")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Now let us have a look at what have happened to the loans with different
purposes. The plot above, shows the distribution of of loan purposes by loan
status.

```{r results='markup'}
df %>%
  filter(!is.na(verification_status))%>%
  group_by(verification_status)%>%
  summarise(mean_annual_inc = mean(annual_inc),
            mean_amnt_loan = mean(loan_amnt),
            n=n())%>%
  ungroup()%>%
  arrange(verification_status)
```

I am curious to know what is the average annual income and average loan amount
granted based on the verification status. As we can see for those whose loans 
could not be verified, the average amount of loan is the lowest and if it was
verified they got the highest amount of loan on average which is exactly what 
we can expect.

```{r results='markup'}
ggplot(data = subset(df, !is.na(verification_status)),
       aes(verification_status, fill = loan_status,color = loan_status))+
  geom_bar(position = "fill")+
  labs(title = "Distribution of verification status by Loan Status")+
  scale_y_continuous(labels = percent_format())
```

Based on the plot above, a very interesting observation is that  the highest
percentage of loans with "Charged Off" Status is from those whose verification
status is "Not Verified" and interestingly, those hwo could verify their status, 
have the highest percentage of "Charged off" loans.


```{r results='markup'}
df %>%
  filter(!is.na(grade))%>%
  group_by(grade)%>%
  summarise(mean_annual_inc = mean(annual_inc),
            mean_amnt_loan = mean(loan_amnt),
            n=n())%>%
  ungroup()%>%
  arrange(grade)
```

Let us see what are the average amount of loans granted and the average annual
income based on the loan grades. So the highest average amount of loans granted 
is on Grade G with about 5400 cases which is of course a very low number compared
to say Grade B with more than 250 thousand cases. For grade B with the highest
number of loans granted, the average amount of loan is $13642.

```{r results='markup'}
ggplot(data = subset(df, !is.na(grade)),
       aes(grade, fill = loan_status,color = loan_status))+
  geom_bar(position = "fill")+
  labs(title = "Distribution of Loan Grade by Loan Status")+
  scale_y_continuous(labels = percent_format())
```

another interesting plot is shown above, this distribution plot gives the
percentage of loan status based on their grades. So, as we expect, grade G has
the highest percentage of loans with "Charged off" status and it shows a
steady downtrend as we move towards the better grades (A and B). 
Moreover, the same trend is evident for the late payments. For example the
highest percentage of late payments are among the loans with grade G.

```{r results='markup'}
df %>%
  filter(!is.na(loan_status))%>%
  group_by(loan_status)%>%
  summarise(mean_annual_inc = mean(annual_inc),
            mean_amnt_loan = mean(loan_amnt),
            n=n())%>%
  ungroup()%>%
  arrange(loan_status)
```


## Creating Bins for Interest Rates, Loan Amounts, and Annual Income

```{r results='markup'}
df <- transform(df, int_group=cut(int_rate,
    breaks=c(0, 7.5, 10, 12.5, 15,20),
      labels=c('0-7.5', '7.5-10', '10-12.5', '12.5-15', '15 and above')))

df <- transform(df, loan_amnt_group=cut(loan_amnt,
    breaks=c(0, 5000, 10000, 15000, 20000, 25000, 40000),
      labels=c('0-5000', '5000-10000', '10000-15000', '15000-20000', '20000-25000','25000 and above')))

df <- transform(df, ann_income_group=cut(annual_inc,
    breaks=c(0, 25000, 50000, 75000, 100000, 1000000),
      labels=c('0-25000', '25000-50000', '50000-75000', '75000-100000', '100000 and above')))
```

We just binned the Interest Rate, Loan Amount, and Annual income columns and 
converted them into Factor values.


## Correlation Analysis Using corrplot

```{r results='markup'}
new_list <- numerical_vars[c('loan_amnt', 'int_rate', 'funded_amnt', 'installment')]
M <- cor(new_list)
corrplot(M, method = "color")
```


```{r results='markup'}
corrplot(M, method = "number", col='black')
```

Above, we have provided the correlation plots for 4 numeric columns of:
'loan_amnt', 'int_rate', 'funded_amnt', 'installment'.


**converting Issue_d column to date Format:**

```{r results='markup'}
df$dates <- dmy(paste0("01-", df$issue_d))
df_amount_grade<- df %>%
  group_by(dates, grade)%>%
  summarise(Loan_Amount = sum(loan_amnt), n=n())
```


```{r results='markup'}
ggplot(df_amount_grade, aes(x = dates, y = Loan_Amount))+
  geom_area(aes(fill=grade))+
  xlab("Issue Date")+
  labs(title = "Total $ Amount of loans grouped by Grade")
```

let us investigate how the total amount of loans has changed over the time.
The plot above shows the total dollar amount of loans granted from 2008 untill
2016.There has been some spikes and some very sharp declines that might be due
to some technical problem or lack of data or perhaps some seasonal trend 
exisiting in the market.


```{r results='markup'}
ggplot(data = subset(df, !is.na(addr_state)),
       aes(x = addr_state, y = loan_amnt, fill = addr_state))+
  stat_summary(fun.y="sum", geom="bar")+
  coord_flip()+
  labs(y ="Total Loan Amount",x="State",title="Bar Chart of loan amount for each state")

```

The amount of loans granted by state is presented above. California is by far
the first state in terms of the total amount of loans.
------

# Bivariate and Multivariate Analysis

* **The relationships I have observed**

I think, most of the finding align with what we could expect before running
doing any analysis, such as the fact that the higher the income, the easier it
gets to get a loan, or the cheapest it gets for you to secure a loan. However,
there was one interesting finding here, my observation showed that  the highest
percentage of loans with "Charged Off" Status is from those whose verification
status is "Not Verified" and interestingly, those hwo could verify their status, 
have the highest percentage of "Charged off" loans.


------

# Final Plots and Summary


### Plot One
```{r results='markup', Plot_One}
ggplot(df_amount_grade, aes(x = dates, y = Loan_Amount))+
  geom_area(aes(fill=grade))+
  xlab("Issue Date")+
  labs(title = "Total $ Amount of loans grouped by Grade")
```

### Description One

The Total amount of loans has grown significantly since 2010, there was a 
period of stagnation from 2008 to 2010 before it took some momentum. Grade B and 
C are the most common loan grades and have become more so after 2015. There are 
a few unusual spikes folowed by sharp declines in late 2014 that need a more
comprehensive analysis to figure out the reason behind it. All in all the growth
looks impressive and there does not seem to be any sign of weakness.

### Plot Two
```{r results='markup', Plot_Two}
ggplot(data = subset(df, !is.na(int_rate)), aes(int_rate, color = grade))+
  geom_histogram(bins = 50)+
  facet_grid(grade ~ .)+
  labs(title = "Interest Rate by Loan Grade")
```

### Description Two
The plot shows the disctribution of Interest Rate by Loan Grade. It is kind of
obvious that there are differences between grades in terms of interest rates.
The grades A-C have significantly lower interest rates.

### Plot Three
```{r results='markup', Plot_Three}
ggplot(data = subset(df, !is.na(verification_status)),
       aes(verification_status, fill = loan_status,color = loan_status))+
  geom_bar(position = "fill")+
  labs(title = "Distribution of verification status by Loan Status")+
  scale_y_continuous(labels = percent_format())
```

### Description Three

As I previously mentioned, this plot shows a rather interesting observation.
The "Verified" category has shown a higher percentage of "Charged off" loans 
compared to the other 2 categories. What we would have expected before any deep
analysis, could be the verified loans have a lower risk of default, but it does 
not seem to be the case. Another important fact from this chart that I would
like to point out is that the verified category has higher percentage of "LATE"s
as well which is also counter-intuitive.

# Reflection

For this project I have used the Lending Club Loan Data. We downloaded the
the dataset from Kaggle.com at this link:
(https://www.kaggle.com/wendykan/lending-club-loan-data). This dataset contains
complete loan data for all loans issued through 2007-2015,including the current
loan status (Current, Late, Fully Paid, etc.) and latest payment information. 
The file is a matrix of about 890 thousand observations and 75 variables. 
I worked on this dataset and wrangled and partialy cleaned the data, then using
great R functionalities and GGplot library I created some univariate analyses,
some multi- and bi-variate analyses of the existing variables.
Some of the findings were aligned with our initial expectations and some were a 
little strange.
I believe there are some very interesting insights that we could draw from the
dataset by working extensively on the latest payments, installments, and also 
fees. The other very interesting topic could be analysing the description of the
loan applications and see what are the most successful terms we could use to 
secure a loan. Another aspect that could be focused on is the job titles and
investigate whether there is any significant pattern in the dataset.

=======